A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-14 tidyindex.qmd
a table listing all the notations + visual mode reformat

  <!-- - Why index is useful, why people care about indexes -->
  <!-- - Define what is an index, what is not -->
  <!-- - What is the challenges with current index construction -->
  <!-- - what can be done if people adopt this pipeline/ why it is beneficial? -->
  <!-- - who would benefit from this paper -->

@@

Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal data,

\begin{equation}
\begin{aligned}
\mathbf{x}(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^P(s;t)] \\
& p = 1, 2, \cdots P, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{aligned}
\end{equation}

where $x^l$ is the $p$th variable, $s$ is the spatial location, and $t$ is the temporal order. <!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.--> The spatial space, $D_s \subseteq \mathbb{R}^d$, is generic to include different spatial representations, including a 1D set of $n$ locations $\{D_s: s_1, s_2, \cdots, s_n\}$, a 2D set of longitude and latitude pairs, and $d$-D coordinate set. The higher dimension in the temporal space, $D_t \subseteq \mathbb{R}^m$, can be used to represent for a $m$-D deconstruction of time. For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.  To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order. When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$. Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed.

@@

`Input`: $x_s(t)$,

`Output`:  $x^l_s(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

@@

\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes.  For example,  Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of

@@

\begin{equation}
x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. 
An example of spatial processing is to align variables collected in different resolutions.
When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$. This is a spatial aggregation and if aggregate using the mean, it can be written as

\begin{equation}
g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
\end{equation}

where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the  coarser grid. Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.

@@

\begin{equation}
h_{\tau}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$. Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.

@@

\begin{equation}
[x(s_i;t_j) - \alpha]/\gamma.
\end{equation}

For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.

<!-- Take one or multiple variables to create a new variable,  -->
<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->

@@

\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}

where $p = 1, 2, \cdots, P$. The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process. For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).

@@

<!-- \begin{equation} -->
<!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->
<!-- \end{equation} -->

@@

\begin{equation}
F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
\end{equation}

where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$. A distribution fit typically aims at finding the distribution that best fits the data. 
Analysts may start from a pool of candidate distributions with a chosen fitting method and goodness of fit measure. 
While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has an uncertainty associated with the data fed and the parameter chosen. A reasonable alternative is to understand how much the index values can vary given different distributions, fitting methods, and goodness of fit tests, and whether these variations are negligible in a given application.

@@

`Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
Also, the use of a normal scale needs to be interpreted with caution. 
@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

\begin{equation}
\begin{cases}
c_0 & C_1 \leq x^l(s; t) < C_0 \\
c_1 & C_2 \leq x^l(s; t) < C_1 \\
c_2 & C_3 \leq x^l(s; t) < C_2 \\
\cdots \\
c_z & C_z \leq x^l(s; t)
\end{cases}
\end{equation}

where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category. In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$. In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.

@@

<!-- # 1. unpack or the argumebt s  -->
<!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->
<!-- # 3. evalute  -->
<!-- # 4. update the data object  -->

@@

The function `compute_indexes()` can be used to collectively compute multiple indexes. The tidyindex offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function. For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise). Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs. The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)). The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.

@@

````
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
````

@@

To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour [TODO: reference to local tour] that varies the weight of each dimension slightly to observe how index value and country ranking changes. The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time. A set of countries are selected for this exercise, which includes 1) the top four with GGGI > 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI < 0.6. Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at  https://vimeo.com/847874016.

@@

* Bangladesh 
* when decreasing the weight on politics, difference between countries narrower. 
* increase politics -> all the country has lower index value

>>>>

<!-- - Why index is useful, why people care about indexes -->

<!-- - Define what is an index, what is not -->

<!-- - What is the challenges with current index construction -->

<!-- - what can be done if people adopt this pipeline/ why it is beneficial? -->

<!-- - who would benefit from this paper -->

@@

Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal process,

```{=tex}
\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}
```
where:

-   $x$ is the variable in $\mathbb{R}^p$,

-   $s$ represents the geographic locations in the space $D_s \subseteq \mathbb{R}^m$. This includes a collection of countries of interest and paired longitude and latitude coordinates of locations, and

-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time decomposed into several components, such as year, month, quarter, and season.

<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.-->

<!--For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$. .-->

<!-- To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order. When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$. Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed. -->

| No. | Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
|-----|-----------------|----------|--------------------|---------------------|
| 1   | Temporal processing     | $x(s;t)$          | aaaaa                                                                                                                                              | $x(s; t^\prime)$                                                    |
| 2   | Spatial processing      | $x(s;t)$          | aaaaa                                                                                                                                              |  $x(s^\prime; t)$                                                   |
| 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????                                                       |
| 4   | Scaling                 | $x(s; t)$         | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
| 6   | Distribution fit        | $x(s; t)$         | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1 |
| 7   | Normalising             | $x(s; t)$         | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                     |
| 8   | Benchmarking            | $x(s; t)$         |                                                                                                                                                    | $b(s;t)$                                                            |
| 9   | Simplification          | $x(s; t)$         | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                              |

: An notation overview of the input, operation, and output of each pipeline module.

Some notes:

-   still $x(s,t)$ notation because for all the other modules, $s$ and $t$ are all fixed, then $x(s, t)$ becomes $x$????
-   aaaaa: just a general function over the input????
-   bbbbb: input and output has the same distributional shape
-   ccccc: latex issue

@@

`Input`: $x(t)$, time series of a single variable with fixed locations

`Output`: $x(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

@@

```{=tex}
\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes. For example, Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of

@@

```{=tex}
\begin{equation}
x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. An example of spatial processing is to align variables collected in different resolutions. When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$. This is a spatial aggregation and if aggregate using the mean, it can be written as

```{=tex}
\begin{equation}
g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
\end{equation}
```
where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid. Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.

@@

```{=tex}
\begin{equation}
h_{\tau}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$. Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.

@@

```{=tex}
\begin{equation}
[x(s_i;t_j) - \alpha]/\gamma.
\end{equation}
```
For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$. @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.

<!-- Take one or multiple variables to create a new variable,  -->

<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->

@@

```{=tex}
\begin{equation}
x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
\end{equation}
```
where $p = 1, 2, \cdots, P$. The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process. For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).

@@

<!-- \begin{equation} -->

<!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->

<!-- \end{equation} -->

@@

```{=tex}
\begin{equation}
F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
\end{equation}
```
where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$. A distribution fit typically aims at finding the distribution that best fits the data. Analysts may start from a pool of candidate distributions with a chosen fitting method and goodness of fit measure. While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has an uncertainty associated with the data fed and the parameter chosen. A reasonable alternative is to understand how much the index values can vary given different distributions, fitting methods, and goodness of fit tests, and whether these variations are negligible in a given application.

@@

`Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions. For example, a normal scale, \[0, 1\], or \[0, 100\] may be favored for reporting certain indexes. In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. Also, the use of a normal scale needs to be interpreted with caution. @fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence. This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.

@@

```{=tex}
\begin{equation}
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}
```
where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category. In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$. In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.

@@

<!-- # 1. unpack or the argumebt s  -->

<!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->

<!-- # 3. evalute  -->

<!-- # 4. update the data object  -->

@@

The function `compute_indexes()` can be used to collectively compute multiple indexes. The tidyindex offers wrapper functions, with the prefix idx\_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function. For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise). Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs. The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)). The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.

@@

    gggi %>% 
      init(id = country) %>%
      add_meta(gggi_weights, var_col = variable) %>% 
      dimension_reduction(
        index_new = aggregate_linear(
          ~labour_force_participation:years_with_female_head_of_state,
          weight = weight))

@@

To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour \[TODO: reference to local tour\] that varies the weight of each dimension slightly to observe how index value and country ranking changes. The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time. A set of countries are selected for this exercise, which includes 1) the top four with GGGI \> 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI \< 0.6. Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at https://vimeo.com/847874016.

@@

-   Bangladesh
-   when decreasing the weight on politics, difference between countries narrower.
-   increase politics -\> all the country has lower index value

### 2023-08-14 tidyindex.tex
a table listing all the notations + visual mode reformat

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

Now, we introduce the notation used for describing pipeline modules.
Consider multivariate spatio-temporal data,

\begin{equation}
\begin{aligned}
\mathbf{x}(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^P(s;t)] \\
& p = 1, 2, \cdots P, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{aligned}
\end{equation}

where \(x^l\) is the \(p\)th variable, \(s\) is the spatial location,
and \(t\) is the temporal order. The spatial space,
\(D_s \subseteq \mathbb{R}^d\), is generic to include different spatial
representations, including a 1D set of \(n\) locations
\(\{D_s: s_1, s_2, \cdots, s_n\}\), a 2D set of longitude and latitude
pairs, and \(d\)-D coordinate set. The higher dimension in the temporal
space, \(D_t \subseteq \mathbb{R}^m\), can be used to represent for a
\(m\)-D deconstruction of time. For example, a year-month combination,
\((Y, M)^\prime\), can be used to record monthly precipitation series
from January 1990 to December 2020, where \(Y = 1990, \cdots, 2020\) and
\(M = 1, \cdots, 12\). To simplify the notation, a subscript letter
\(s\) or \(t\) is used to denote the data have a fixed spatial locations
or temporal order. When the spatial locations are fixed, the data become
multivariate time series, denoted as \(x^l_s(t)\). Similarly,
\(x^l_t(s)\) represents multivariate spatial data when the temporal
dimension is fixed.

@@

\texttt{Input}: \(x_s(t)\),

\texttt{Output}: \(x^l_s(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

@@

\begin{equation}
\begin{cases}
c_0 & C_1 \leq x^l(s; t) < C_0 \\
c_1 & C_2 \leq x^l(s; t) < C_1 \\
c_2 & C_3 \leq x^l(s; t) < C_2 \\
\cdots \\
c_z & C_z \leq x^l(s; t)
\end{cases}
\end{equation}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Now, we introduce the notation used for describing pipeline modules.
Consider multivariate spatio-temporal process,

\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{equation}

where:

\begin{itemize}
\item
  \(x\) is the variable in \(\mathbb{R}^p\),
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^d\). This includes a collection of
  countries of interest and paired longitude and latitude coordinates of
  locations, and
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^m\).
  Sometimes, it can be time decomposed into several components, such as
  year, month, quarter, and season.
\end{itemize}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0685}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1370}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2740}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2877}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(s;t)\) & aaaaa & \(x(s; t^\prime)\) \\
2 & Spatial processing & \(x(s;t)\) & aaaaa & \(x(s^\prime; t)\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

Some notes:

\begin{itemize}
\tightlist
\item
  still \(x(s,t)\) notation because for all the other modules, \(s\) and
  \(t\) are all fixed, then \(x(s, t)\) becomes \(x\)????
\item
  aaaaa: just a general function over the input????
\item
  bbbbb: input and output has the same distributional shape
\item
  ccccc: latex issue
\end{itemize}

@@

\texttt{Input}: \(x(t)\), time series of a single variable with fixed
locations

\texttt{Output}: \(x(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

@@

\begin{equation}
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}
